---
title: "Task 4 - Halloween Candy Data"
subtitle: Dirty Data Project
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---
# Halloween Candy Data

## Introduction

The source of the data is

We have three spreadsheet files in our `raw_data` folder.

1. boing-boing-candy-2015.xlsx
2. boing-boing-candy-2016.xlsx
3. boing-boing-candy-2015.xlsx

## Cleaning

### Candy names
In the original dataset there are a great number of columns for each name of 
candy. 

The column names from each spreadsheet were obtained and documented in [`halloween_candy_data_files.html`](halloween_candy_data_files.html). This enabled comparison between the structure
of each sheet. 2015 and 2016 use a similar naming convention with potential 
candy names inside square brackets `[]`, in 2017 this switches to a 
prefix of "Q6 | "

It quickly becomes apparent that the candy names apart from brand names contain
items that are not edible e.g. DVDs. The prompts are often light-hearted and 
this is clearly supposed to be a fun exercise. For this reason candy it was 
decided to leave all candy names in the dataset.

1. It's taking a looser definition of candy as being things that are usually
pleasurable
2. Brand names appear to originate from the USA and many are unfamiliar in the 
UK. In the interests of time we are not researching each candy name column to
see if it is confectionery or not.

### Dataset for analysis

The data from all three spreadsheets has been combined into one CSV file 
`clean_data/halloween_candy.csv`. This is file is created by running script 
`data_cleaning_scripts/cleaning.R`, this step only needs to be done once.

## Analysis questions

```{r}
library(here)
library(tidyverse)
```

```{r}
# read cleaned Halloween candy dataset
# This is the combined data for all three years
halloween_candy <- read_csv(here::here("clean_data/halloween_candy.csv"))

# We have the following columns in the clean data
# age, trick_or_treating, gender, year, country, candy_name, candy_rating, candy_popularity
```


### Q1 Total candy ratings
What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)

```{r}
halloween_candy %>%
  select(candy_rating) %>%
  filter(!is.na(candy_rating)) %>%
  summarise(number_candy_ratings = n())
```

There are **772,352** candy ratings given across the three years.

### Q2 Average age trick or treating

What was the average age of people who are going out trick or treating?

```{r}
halloween_candy %>%
  select(age, trick_or_treating) %>%
  filter(trick_or_treating == "Yes") %>%
  group_by(trick_or_treating) %>%
  summarise(average_age_trick_or_treating = mean(age, na.rm = TRUE)) %>%
  ungroup()
```

Average age of people who are going out trick or treating is **35.38839**.

### Q3 Average age not trick or treating

What was the average age of people who are not going trick or treating?

```{r}
halloween_candy %>%
  select(age, trick_or_treating) %>%
  filter(trick_or_treating == "No") %>%
  group_by(trick_or_treating) %>%
  summarise(average_age_not_trick_or_treating = mean(age, na.rm = TRUE)) %>%
  ungroup()
```

Average age of people not going trick or treating is **39.25324
**.  
Assumption: we are only calculating results on those who explicitly answered "No" rather than also include those who did not answer the question.

### Q4 Most ratings for joy, despair and meh

For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}
halloween_candy %>%
  select(candy_rating, candy_name) %>%
  group_by(candy_rating, candy_name) %>%
  summarise(number_of_ratings = n()) %>%
  slice_max(number_of_ratings, n = 1) %>%
  arrange(desc(number_of_ratings)) %>%
  ungroup()
```

* For joy, `Any full-sized candy bar` received most ratings.
* For despair, `Broken glow stick` received most ratings.
* For meh, `Lollipops` received most ratings.

### Q5 Starburst rated as despair

How many people rated Starburst as despair?

```{r}
halloween_candy %>%
  select(candy_name, candy_rating) %>%
  filter(candy_name == "Starburst" & candy_rating == "DESPAIR") %>%
  group_by(candy_rating) %>%
  summarise(number_ratings_starburst_despair = n()) %>%
  ungroup()
```

**1990** people rated Starburst as despair?

***
For the next three questions, count despair as -1, joy as +1, and meh as 0.

***
### Q6 Popular candy bar by gender

What was the most popular candy bar by this rating system for each gender in the dataset ?

```{r}
halloween_candy %>%
  select(candy_popularity, candy_name, gender) %>%
  filter(!is.na(gender)) %>%
  group_by(gender, candy_name) %>%
  summarise(total_popularity = sum(candy_popularity, na.rm = TRUE)) %>%
  slice_max(total_popularity, n = 1) %>%
  arrange(desc(total_popularity)) %>%
  ungroup()
```

* Most popular candy bar for `Male` was **Any full-sized candy bar**.
* Most popular candy bar for `Female` was **Any full-sized candy bar**.
* Most popular candy bar for `I'd rather not say` was **Any full-sized candy bar**.
* Most popular candy bar for `Other` was **Any full-sized candy bar**.

### Q7 Popular candy bar by year

What was the most popular candy bar in each year?

```{r}
```

### Q8 Popular candy bar by country

What was the most popular candy bar by this rating for people in US, Canada, UK,
and all other countries?

```{r}
```
