---
title: "Task 4 - Halloween Candy Data"
subtitle: Dirty Data Project
author: "Lesley Duff"
date: "`r format(Sys.Date())`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---
# Halloween Candy Data

## Introduction

We have data from a light-hearted online 'CANDY HIERARCHY' survey conducted from
2014 to 2017 by [Dave Ng of the University of British Columbia](https://www.msl.ubc.ca/people/dr-dave-ng/). The main aim of the survey
was to gather what users feel about different types of candy by giving them 
ratings.

The survey also asks about whether users are participating in the tradition of 
[trick or treating](https://en.wikipedia.org/wiki/Trick-or-treating) - better 
known as 'guising' in Scotland.

> "The "treat" is some form of confectionery, usually candy/sweets, although in 
some cultures money is given instead." [Wikipedia]

The results were analysed and published on
the Boing Boing website e.g. you can download a poster of a visual analysis of 
the most recent year (2017).

* [visualisation of the results from 2017 (PDF)](https://boingboing.net/wp-content/uploads/2017/10/candyhierarchy2017.pdf)

The survey data for a number of years is available for download.

**Data Source:** [So Much Candy Data, Seriously](https://www.scq.ubc.ca/so-much-candy-data-seriously/)

## Assumptions

### Non-technical
* The tone of the survey is light-hearted - it's categorised on the Boing Boing 
website as

>creative / humour / math / web experiment

* The survey questions are a mixture of fixed 'candy' items but also 
free text for members of the general public from around the world. 
The datasets produced are website user generated content, a mixture of answers 
to structured survey questions but also *unstructured* and *unfiltered* text. 
Users themselves have responded in humorous ways. But the survey was still 
conducted in a 'formal' way gathering responses via web forms.

* Some ages provided are for young children, parents/adults are possibly 
filling out the form on behalf of their child.

## Technical

We have three spreadsheet files representing survey results from 2015, 2016 and 
2017 in our `raw_data` folder.  
N.B. These file names have been standardised from what you will find on the 
data source website.

1. [boing-boing-candy-2015.xlsx](../raw_data/boing-boing-candy-2015.xlsx)
2. [boing-boing-candy-2016.xlsx](../raw_data/boing-boing-candy-2016.xlsx)
3. [boing-boing-candy-2017.xlsx](../raw_data/boing-boing-candy-2017.xlsx)


* The data source website contains open data from 2014-2017 however we will not 
be including 2014 as

>Unfortunate quirks in the data include the fact that the 2014 data is not the 
raw set (can’t seem to find it)"

* For analysis efficiency the cleaning done for analysis will only create a new
dataset containing **only** the items *necessary* for the analysis, other 
columns will be ignored. This restructuring comes at the expense of increasing 
the file sizes involved. e,g, original `raw_data` < 4MB whereas the `clean_data` file for analysis is ~35MB.

* The columns vary from year to year but our cleaned data will standardise the 
naming so all three years of data will be combined in one dataset.

## Cleaning

### Candy names

In the original dataset there are a great number of columns for each name of 
candy. 

The column names from each spreadsheet were obtained and documented in [`halloween_candy_data_files.html`](halloween_candy_data_files.html). This 
enabled comparison between the structure of each sheet. 2015 and 2016 use a 
similar naming convention with potential candy names inside square brackets 
`[]`, in 2017 this switches to a prefix of `"Q6 | "`

It quickly becomes apparent that the candy names apart from brand names contain
items that are not edible e.g. DVDs. The prompts are often light-hearted and 
this is clearly supposed to be a fun exercise. For this reason it was 
decided to leave all original candy names in the dataset.

1. It's taking a broader definition of 'candy' as being things that are usually
pleasurable not just confectionery
2. Brand names appear to originate from North America and many are unfamiliar 
in the UK. In the interests of time we are not researching each candy name 
column to see if it is confectionery or not.

In a similar vein there are a couple of columns left as free text entry for the
survey users e.g. "Please list any items not included above that give you JOY" 
and for despair. These are a potential source of extra ratings. These also 
contain sex and drugs references - such columns have been omitted from this 
analysis.

### Country names

Question 8 requires country names and asks specifically for **US, Canada, UK** by name. Countries were not asked for in the 2015 survey so for that year in the new dataset have been set to NA. For 2016/2017 there are a number of issues Particularly for US many references are jokey as if being chanted. Or slang 
versions like `murica`
Some users have entered the name of US states instead of their country.
For Canada there are some punctuation typos
For UK there are misspellings of United Kingdom and individual countries such as Scotland and England.

To try to address some of the common variations found a helper function 
`clean_country()` has been added `data_cleaning_scripts/clean_country.R`. This is
called from the main coordinating cleaning script in places where user-supplied 
countries have been provided. Adding search and replace features like these 
leads to a slightly longer amount of time taken for the initial clean but makes 
for a simpler analysis.

### Ages

This was a character column and especially in earlier years has a few values as 
jokey replies to the question. It was decided to set non numeric values to NA 
and then these text observations are dropped from analysis.
N.B. This conversion may generate warning messages "NAs introduced by coercion".

The numerical age provided was checked to see if it was in a reasonable range.
As of 2023-01-02 [oldest known individual who ever lived](https://en.wikipedia.org/wiki/List_of_the_verified_oldest_people) was 122
We also excluded 0 or entries for children under 3. We've left children 3 and 
above in on the assumption that parents are possibly completing the form on 
behalf of their child and trick or treating usually involves walking to 
houses. If age values were outside the chosen range they were changed to NA and 
not included in analysis.

### Dataset for analysis

A cleaning script has been created. This is used by running script 
`data_cleaning_scripts/cleaning.R`
After the script completes the data from all three spreadsheets is been combined 
into one large CSV file `clean_data/halloween_candy.csv`. This step only needs 
to be done once as the analysis will use the cleaned data.

### Analysis preparation

The eight questions to be answered were examined to suggest which columns from 
the original spreadsheet would be required as a minimum to enable results to be 
generated. This analysis preparation documentation can be found at 
[cleaning_columns_required.html](cleaning_columns_required.html).

```{r}
library(here)
library(tidyverse)
```

```{r}
# read cleaned Halloween candy dataset
# This is the combined data for all three years
halloween_candy <- read_csv(here::here("clean_data/halloween_candy.csv"))
```
We have the following columns in the clean data

* `age`
* `trick_or_treating`
* `gender`
* `year`
* `country`
* `candy_name`
* `candy_rating`
* `candy_popularity`


## Analysis questions


### Q1 Total candy ratings
What is the total number of candy ratings given across the three years. (Number of candy ratings, not the number of raters. Don’t count missing values)

```{r}
halloween_candy %>%
  select(candy_rating) %>%
  filter(!is.na(candy_rating)) %>%
  summarise(number_candy_ratings = n())
```

There are **772,352** candy ratings given across the three years.

### Q2 Average age trick or treating

What was the average age of people who are going out trick or treating?

```{r}
halloween_candy %>%
  select(age, trick_or_treating) %>%
  filter(trick_or_treating == "Yes") %>%
  group_by(trick_or_treating) %>%
  summarise(average_age_trick_or_treating = mean(age, na.rm = TRUE)) %>%
  ungroup()
```

Average age of people who are going out trick or treating is **35.38839**.

### Q3 Average age not trick or treating

What was the average age of people who are not going trick or treating?

```{r}
halloween_candy %>%
  select(age, trick_or_treating) %>%
  filter(trick_or_treating == "No") %>%
  group_by(trick_or_treating) %>%
  summarise(average_age_not_trick_or_treating = mean(age, na.rm = TRUE)) %>%
  ungroup()
```

Average age of people not going trick or treating is **39.25324
**.  
Assumption: we are only calculating results on those who explicitly answered "No" rather than also include those who did not answer the question.

### Q4 Most ratings for joy, despair and meh

For each of joy, despair and meh, which candy bar received the most of these ratings?

```{r}
halloween_candy %>%
  select(candy_rating, candy_name) %>%
  group_by(candy_rating, candy_name) %>%
  summarise(number_of_ratings = n()) %>%
  slice_max(number_of_ratings, n = 1) %>%
  arrange(desc(number_of_ratings)) %>%
  ungroup()
```

* For joy, `Any full-sized candy bar` received most ratings.
* For despair, `Broken glow stick` received most ratings.
* For meh, `Lollipops` received most ratings.

### Q5 Starburst rated as despair

How many people rated Starburst as despair?

```{r}
halloween_candy %>%
  select(candy_name, candy_rating) %>%
  filter(candy_name == "Starburst" & candy_rating == "DESPAIR") %>%
  group_by(candy_rating) %>%
  summarise(number_ratings_starburst_despair = n()) %>%
  ungroup()
```

**1990** people rated Starburst as despair?

***
For the next three questions, count despair as -1, joy as +1, and meh as 0.

***
### Q6 Popular candy bar by gender

What was the most popular candy bar by this rating system for each gender in the dataset ?

```{r}
halloween_candy %>%
  select(candy_popularity, candy_name, gender) %>%
  filter(!is.na(gender)) %>%
  group_by(gender, candy_name) %>%
  summarise(total_popularity = sum(candy_popularity, na.rm = TRUE)) %>%
  slice_max(total_popularity, n = 1) %>%
  arrange(desc(total_popularity)) %>%
  ungroup()
```

* Most popular candy bar for `Male` was **Any full-sized candy bar**.
* Most popular candy bar for `Female` was **Any full-sized candy bar**.
* Most popular candy bar for `I'd rather not say` was **Any full-sized candy bar**.
* Most popular candy bar for `Other` was **Any full-sized candy bar**.

Assumption: we are not including ratings where no response to the gender 
question was given.

### Q7 Popular candy bar by year

What was the most popular candy bar in each year?

```{r}
halloween_candy %>%
  select(candy_popularity, candy_name, year) %>%
  group_by(year, candy_name) %>%
  summarise(total_popularity = sum(candy_popularity, na.rm = TRUE)) %>%
  slice_max(total_popularity, n = 1) %>%
  arrange(year) %>%
  ungroup()
```

* In 2015 the most popular candy bar was `Any full-sized candy bar`.
* In 2016 the most popular candy bar was `Any full-sized candy bar`.
* In 2017 the most popular candy bar was `Any full-sized candy bar`.

### Q8 Popular candy bar by country

```{r eval=FALSE}
# This code was used during development to view country names
# From the names returned patterns were added to the clean_country function
distinct_countries <- halloween_candy %>%
  select(country) %>%
  filter(!is.na(country)) %>%
  distinct() %>%
  arrange(country)

distinct_countries
```
What was the most popular candy bar by this rating for people in US, Canada, UK,
and all other countries?

```{r}
# List of countries referenced by name in question 8
main_countries <- c("US", "Canada", "UK")

halloween_candy %>%
  select(candy_popularity, candy_name, country) %>%
  filter(!is.na(country)) %>%
  mutate(country = if_else(country %in% main_countries,
    country, "Other"
  )) %>%
  group_by(country, candy_name) %>%
  summarise(total_popularity = sum(candy_popularity, na.rm = TRUE)) %>%
  slice_max(total_popularity, n = 1) %>%
  arrange(desc(total_popularity)) %>% 
  ungroup()
```

* Most popular candy bar by this rating US, `Any full-sized candy bar`
* Most popular candy bar by this rating Canada, `Any full-sized candy bar`
* Most popular candy bar by this rating UK, `Cash, or other forms of legal tender`
* Most popular candy bar by this rating all other countries, `Any full-sized candy bar`

Assumption: 2015 data is excluded from these rseults as country wasn't something
asked for then. 

A bit of home interest - UK folk buck the rest of the world in choosing money 
over sweeties...and this was *before* the cost of living crisis.

## Other analysis

### Top ten popular candy in UK

What were the top ten popular candy by this rating for people in UK?

```{r}
halloween_candy %>%
  select(candy_popularity, candy_name, country) %>%
  filter(!is.na(country) & country == "UK") %>%
  group_by(candy_name) %>%
  summarise(total_popularity = sum(candy_popularity, na.rm = TRUE)) %>%
  slice_max(total_popularity, n = 10) %>% 
  arrange(desc(total_popularity), candy_name) %>% 
  ungroup()
```

