---
title: "Task 1 - Decathlon Data"
subtitle: "Dirty Data Project"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
date: "`r format(Sys.Date())`"
author: "Lesley Duff"
---
# Decathlon Data

## Introduction
We have a dataset `raw_data/decathlon.rds` relating to a 
[Decathlon](https://en.wikipedia.org/wiki/Decathlon) event.

> "The decathlon is a combined event in athletics consisting of ten track and 
field events...Performance is judged on a points system in each event, not by 
the position achieved"

There are ten events over two days:

* 100 metres
* Long jump
* Shot put
* High jump
* 400 metres
* 110 metres hurdles
* Discus throw
* Pole vault
* Javelin throw
* 1500 metres

We have details of the results by each athlete for each of the ten events. 
Also available is the athlete `rank` and `points` in each `competition`. 

There are 32 different athletes in the results.

The source of the data is [decathlon: Performance in decathlon (data)](https://rdrr.io/github/husson/FactoMineR/man/decathlon.html). This 
source claims the details are from events in 2004.

## Assumptions

### Technical

* A script `data_cleaning_scripts/cleaning.R` has been run prior to 
analysis. The original dataset file `raw_data/decathlon.rds`  cleaned
dataset as a new CSV file `clean_data/decathlon.csv`. CSV is a more accessible 
format usable by a wider range of software. All further analysis was done using 
the new CSV file.

### Non-technical

* Due to the order and number of the columns it is assumed that this data is 
a male decathlon. Females compete in this event more rarely and have a 
different order of decathlon events.
* Preliminary cleaning of data reveals we have data for two separate 
competitions - `Decastar` and `OlympicG` (Olympic Games).

## Cleaning

The following changes to the dataset were made by the cleaning script `data_cleaning_scripts/cleaning.R`:

* The column titles were renamed to remove full stops and start columns with 
numbers with an 'x' using `janitor::clean_names()`
* Columns `javeline` and `x110m_hurdle` were renamed to `javelin` and 
`x110m_hurdles` to improve the match to the English descriptions of the names of
the events.
* The original data had row names with the names of the athletes. These were
moved to be a new column `athlete`.

## Analysis questions   

```{r}
library(here)
library(tidyverse)
```

```{r}
# read cleaned decathlon dataset
decathlon_data <- read_csv(here::here("clean_data/decathlon.csv"))
```

### Who had the longest long jump seen in the data?

```{r}
decathlon_data %>%
  select(athlete, decathlon_event, result) %>%
  filter(decathlon_event == "long jump") %>%
  slice_max(result, n = 1)
```

**Clay** had the longest long jump at **7.96m**.

### What was the average 100m time in each competition?

```{r}
decathlon_data %>%
  select(decathlon_event, competition, result) %>%
  filter(decathlon_event == "100m") %>%
  group_by(competition) %>%
  summarise(average_100m_time = mean(result, na.rm = TRUE))
```

The average 100m time in **Decastar** was **11.17538s**.  
The average 100m time in **Olympic Games** was **10.91571s**.

### Who had the highest total points across both competitions?

```{r}
decathlon_data %>%
  select(athlete, points, competition) %>%
  # lots of duplicates, get rid of them
  unique() %>%
  group_by(athlete) %>%
  summarise(highest_total_points = sum(points, na.rm = TRUE)) %>%
  slice_max(highest_total_points, n = 1)
```
**Sebrle** has the highest total points across both competitions, **17,110**.

This is assuming that we are totaling the points obtained by each athlete in all
competitions they participated in. Some athletes only participated in one 
competition.

### What was the shot-put scores for the top three competitors in each competition?

```{r}
decathlon_data %>%
  select(decathlon_event, score = result, athlete, competition, rank, points) %>%
  filter(decathlon_event == "shot put") %>%
  group_by(competition) %>%
  slice_max(points, n = 3)
```

In **Decastar** shot-put scores for the top three competitors were 

1. 14.83m - Sebrle
2. 14.26m - Clay
3. 14.77m - Karpov	

In **Olympic Games** shot-put scores for the top three competitors were 

1. 16.36 - Sebrle
2. 15.23 - Clay
3. 15.93 - Karpov

This is assuming that we are measuring 'top' as highest points by a competitor 
in a competition.

### What was the average points for competitors who ran the 400m in less than 50 seconds vs. those than ran 400m in more than 50 seconds?

```{r}
decathlon_data %>%
  select(points, decathlon_event, result) %>%
  filter(decathlon_event == "400m" & result < 50) %>%
  summarise(average_points_x400m_under_50s = mean(points, na.rm = TRUE))

decathlon_data %>%
  select(points, decathlon_event, result) %>%
  filter(decathlon_event == "400m" & result > 50) %>%
  summarise(average_points_x400m_over_50s = mean(points, na.rm = TRUE))
```
average points for competitors who ran the 400m in less than 50 seconds **8120.483**  
vs.  
average points for competitors who ran 400m in more than 50 seconds **7727.167**

## Other observations

### How many athletes took part in each competition?

```{r}
decathlon_data %>%
  select(athlete, competition) %>%
  # remove duplicates
  unique() %>%
  group_by(competition) %>%
  summarise(num_athletes = n())
```
**13** athletes took part in **Decastar**.  
**28** athletes took part in **Olympic Games**.

### Which athletes took part in both competitions?

```{r}
decathlon_data %>%
  select(athlete, competition) %>%
  unique() %>%
  group_by(athlete) %>%
  summarise(number_competitions = n()) %>%
  filter(number_competitions == 2) %>%
  arrange(athlete)
```
The following athletes took part in both competitions:

**Barras, Bernard, Clay, Hernu, Karpov, Nool, Sebrle, Warners, Zsivoczky**

### Results by event

The best result of each event across both competitions are as follows.

```{r}
# 100m
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "100m") %>%
  summarise(best_100m = min(result))

# long jump
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "long jump") %>%
  summarise(best_long_jump = max(result))

# shot put
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "shot put") %>%
  summarise(best_shot_put = max(result))

# high jump
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "high jump") %>%
  summarise(best_high_jump = max(result))

# 400m
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "400m") %>%
  summarise(best_400m = min(result))

# 110m hurdles
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "110m hurdles") %>%
  summarise(best_110m_hurdles = min(result))

# discus
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "discus") %>%
  summarise(best_discus = max(result))

# pole vault
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "pole vault") %>%
  summarise(best_pole_vault = max(result))

# javelin
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "javelin") %>%
  summarise(best_javelin = max(result))

# 1500m
decathlon_data %>%
  select(decathlon_event, result) %>%
  filter(decathlon_event == "1500m") %>%
  summarise(best_1500m = min(result))
```
Best results for each event are as follows:

| Event              | Results | 
| -------------------| -------: |
| 100 metres         | 10.44 |
| Long jump          | 7.96m |
| Shot put           | 16.36m |
| High jump          | 2.15m |
| 400 metres         | 46.81 |
| 110 metres hurdles | 13.97 |
| Discus throw       | 51.65m |
| Pole vault         |  5.4m |
| Javelin throw      | 70.52m	|
| 1500 metres        | 4:22.10 |
