---
title: "Task 1 - Decathlon Data"
subtitle: "Dirty Data Project"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    number_sections: true
date: "`r format(Sys.Date())`"
author: "Lesley Duff"
---
# Decathlon Data

## Introduction
We have a dataset relating to a [Decathlon](https://en.wikipedia.org/wiki/Decathlon) event.

> "The decathlon is a combined event in athletics consisting of ten track and field events...Performance is judged on a points system in each event, not by the position achieved"

There are ten events over two days:

* 100 metres
* Long jump
* Shot put
* High jump
* 400 metres
* 110 metres hurdles
* Discus throw
* Pole vault
* Javelin throw
* 1500 metres

We have details of the athlete `rank` and `points` in each 
`competition`.

There are 32 different athletes in the results.


## Assumptions

* Due to the order and number of the columns I am assuming that this data is 
a male decathlon. Women have a different order of decathlon events and are more 
likely to compete in a seven event 
[heptathlon](https://en.wikipedia.org/wiki/Heptathlon) instead.
* Preliminary cleaning of data reveals we have data for two separate 
competitions - `Decastar` and `OlympicG`.


## Cleaning

* We renamed the column titles to remove full stops and start columns with numbers
with an 'x'
* We renamed javeline and x110m_hurdle to javelin and x110m_hurdles to better 
match the English descriptions of the names of the events.

## Analysis questions   

```{r}
library(here)
library(tidyverse)

# use this to set the path to the data file
decathlon_data <- read_csv(here("clean_data/decathlon.csv"))
```

### Who had the longest long jump seen in the data?

```{r}
decathlon_data %>% 
  select(athlete, long_jump) %>% 
  arrange( desc(long_jump)) %>% 
  slice_max(long_jump, n = 1)
```

Clay had the longest long jump at 7.96m.

### What was the average 100m time in each competition?

```{r}
decathlon_data %>% 
  select(x100m, competition) %>% 
  group_by(competition) %>% 
  summarise(average_100m_time = mean(x100m, na.rm = TRUE))

```

### Who had the highest total points across both competitions?
```{r}
decathlon_data %>% 
  select(athlete, points, competition)  %>% 
  slice_max(points, n = 1)
```


### What was the shot-put scores for the top three competitors in each competition?

```{r}
decathlon_data %>% 
  select(athlete, shot_put, competition) %>% 
  group_by(competition) %>% 
  slice_max(shot_put, n = 3)

```



### What was the average points for competitors who ran the 400m in less than 50 seconds vs. those than ran 400m in more than 50 seconds?
